\section{Introduction}

The Collatz map is defined by the iteration
\[
T(n)=
\begin{cases}
n/2, & n \equiv 0 \pmod 2,\\
3n+1, & n \equiv 1 \pmod 2.
\end{cases}
\]
The conjecture states that every positive integer reaches $1$ under repeated application of $T$.
This statement is easy to explain and easy to compute, yet it remains unresolved.
The tension between elementary definition and elusive global behavior makes Collatz a useful case study for computational mathematics: it supports large-scale experimentation, but it also punishes overconfident interpretation.

This paper takes an explicitly empirical and reproducible stance.
Rather than claiming a proof, we build and document a workflow for generating auditable evidence, stress-testing hypotheses, and producing figures that can be regenerated from code.
Our premise is pragmatic: for problems where formal progress is hard, disciplined computational practice can still sharpen questions, reject weak ideas quickly, and expose robust structural phenomena.

\paragraph{Research goals.}
We focus on three goals:
\begin{itemize}
  \item \textbf{Baseline structure.} Quantify stopping times $\sigma(n)$, peak values $\pi(n)$, and their joint behavior over large finite ranges.
  \item \textbf{Model extensions.} Probe nearby dynamical families, including generalized piecewise maps (for example, rules of the form $n \mapsto n/p$ when divisible by $p$, else $an+c$) and analytic extensions into the complex plane.
  \item \textbf{Interpretability.} Develop visual and graph-based diagnostics that make trajectory organization, residue transitions, and failure modes easier to inspect.
\end{itemize}

\paragraph{Contributions.}
The concrete contributions of this work are:
\begin{itemize}
  \item a reproducible experiment stack linking scripts, datasets, and paper artifacts;
  \item baseline pipelines for range statistics and figure generation;
  \item a residue-class transition framework for comparing classical and generalized rules;
  \item a complex-plane escape-map prototype for exploratory dynamical analysis;
  \item an agent-oriented research workflow for coordinating code, paper writing, and bibliography review.
\end{itemize}

The central theme is methodological clarity: each claim in the manuscript is intended to be traceable to a script, parameter set, and output artifact.
This design reduces ambiguity, improves reviewability, and supports iterative extension by multiple contributors.

\paragraph{Scope and limitations.}
All reported results are finite-window computations.
Observed regularities may be suggestive, but they are not proofs of asymptotic behavior.
Generalized rule scans are included as comparative experiments, not as equivalent formulations of the original Collatz conjecture.

\paragraph{Paper organization.}
Section~\ref{sec:related-work} positions this study within prior literature.
Section~\ref{sec:methods} defines datasets, protocols, and reproducibility constraints.
Section~\ref{sec:experiments} reports baseline and generalized-rule experiments.
Section~\ref{sec:visualizations} presents visual diagnostics, and Section~\ref{sec:new-ideas} records testable hypotheses for subsequent work.
